---
title: "CCTV Stormwater Pipe Project - Data Visualization"
author: "Fan"
date: '2022-03-27'
output: html_document:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.aligh = 'center',
                      echo = TRUE)
```


```{r library, include=FALSE}
library(ggplot2)
library(tidyverse)
library(ggmap)
library(devtools)
library(maps)
library(ggpubr)
library(corrplot)
library(PerformanceAnalytics)
library(caret)
library(ROSE)

# Import dataset
SW_data <- read.csv("C:/Users/fan100199/OneDrive - Mott MacDonald/Desktop/MottMac/Data_sets\\df_clean2.csv")
```

# 1. Geomap 
```{r}
# Coordinates for location of interest
akl_map <- get_stamenmap(
  bbox = c(left = 174.55, bottom = -37.0000, 
           right = 174.95, top = -36.7500),
  maptype = "terrain",
  zoom = 12
)
ggmap(akl_map)

```
# 2. Response variables
```{r}
summary(SW_data)
# Convert values of conditions to 1 and 0,
# 1 means condition grades equal to or higher than 4,
# 0 means condition grades lower than 4. 
SW_data$structural_condition[SW_data$structural_condition <= 3] <- 0
SW_data$structural_condition[SW_data$structural_condition > 3] <- 1

SW_data$service_condition[SW_data$service_condition <= 3] <- 0
SW_data$service_condition[SW_data$service_condition > 3] <- 1

# Convert the type from numberic to factors
SW_data$structural_condition <- as.factor(SW_data$structural_condition)
SW_data$service_condition <- as.factor(SW_data$service_condition)
# Bar plot of pipe conditions
# structural condition
ggplot(SW_data, aes(x=structural_condition, fill = structural_condition)) + 
  geom_bar(alpha=0.6) + 
  scale_fill_discrete(name = "Structual Condition", 
                      labels = c("Good", "Bad")) +
  labs(title = "Distribution of Pipe Structural Condition",
       x = "Structural Condition",
       y = "Count")
# service condition
ggplot(SW_data, aes(x=service_condition, fill = service_condition)) + 
  geom_bar(alpha=0.6) + 
  scale_fill_discrete(name = "SErvice Condition", 
                      labels = c("Good", "Bad")) +
  labs(title = "Distribution of Pipe Service Condition",
       x = "Service Condition",
       y = "Count")


```

# 3. Independent Vairables

## 3.1 Pipe Age

Pipe age was calculated by the difference between pipe installation date and inspection date. 
```{r}
# Visualization plots
# Histogram plot of pipe age
annotations <- data.frame(
  x = c(round(min(SW_data$years), 2), round(mean(SW_data$years), 2), round(max(SW_data$years), 2)),
  y = c(4, 52, 5),
  label = c("Min:", "Mean:", "Max:")
) 
age1 <- ggplot(data=SW_data, aes(x=years)) +
  geom_histogram(color="black", fill="grey", bins = 30, boundary = 0) + 
  geom_text(data = annotations,  aes(x = x, y = y, label = paste(label, x)), size = 3, fontface = "bold") +
  labs(title = "Histogram of Pipe Age", 
       x = "Age(Years)", 
       y = "Count")

age2 <- ggplot(SW_data, aes(x=years, y = factor(structural_condition), fill = factor(structural_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Pipe Age by Structural Condition",
       fill = "Structural Condition",
          y = " ",
          x = "Age") +
  scale_fill_discrete(labels = c("Good", "Poor"))

age3 <- ggplot(SW_data, aes(x=years, y = factor(service_condition), fill = factor(service_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Pipe Age by Service Condition",
       fill = "Service Condition",
          y = " ",
          x = "Age") +
  scale_fill_discrete(labels = c("Good", "Poor"))

ggarrange(age1, # first row with histogram plot
          ggarrange(age2, age3, ncol=2), # second row with two box plots
          nrow = 2)
```

### 3.1.1 Pipe Age Density Plot by conditions
```{r}
# density plot of pipe age by structural condition
ggplot(data=SW_data, aes(x=years, fill=structural_condition)) +
  geom_density(alpha=0.6) +
  facet_grid(structural_condition ~ .) +
  labs(title = "Histogram of Pipe Age by Structural Condition", 
       x = "Age(Years)", 
       y = "Density",
       fill = "Structural Condition") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 10))

# density plot of pipe age by service condition
ggplot(data=SW_data, aes(x=years, fill=service_condition)) +
  geom_density(alpha=0.6) +
  facet_grid(service_condition ~ .) +
  labs(title = "Histogram of Pipe Age by Service Condition", 
       x = "Age(Years)", 
       y = "Density",
       fill = "Service Condition") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 10))
```

## 3.2 Pipe Material 
```{r}
# Bar chart of pipe material
SW_data %>% 
    count(mat = factor(material)) %>% 
    mutate(pct = prop.table(n)) %>% 
    ggplot(aes(x = mat, y = pct, label = scales::percent(pct))) + 
    geom_col(position = 'dodge', fill="grey", color="black") + 
    geom_text(position = position_dodge(width = .9),    # move to center of bars
              vjust = -0.5,    # nudge above top of bar
              size = 3) + 
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Percentage of Pipe Materials",
         x = "Material",
         y = "Percentage")

# Save a table of material and structural condition
table1 <- table(SW_data$structural_condition, SW_data$material)
table2 <- table(SW_data$service_condition, SW_data$material)
# view tables
table1
# Plot the result
barplot(prop.table(table1,margin = 2),
        legend.text = TRUE,
        args.legend = list(x = "topright"),
        beside = TRUE,
        main = "Percentage of Pipes in Good and Poor Structural Condition by Material")
barplot(prop.table(table2,margin = 2),
        legend.text = TRUE,
        args.legend = list(x = "topright"),
        beside = TRUE,
        main = "Percentage of Pipes in Good and Poor Service Condition by Material")

# boxplot of distribution of pipe age by different materials
boxplot(SW_data$years ~ SW_data$material)


```
## 3.3 pipe diameter (mm)
```{r}
# Histogram plot
h_pd <- ggplot(SW_data, aes(diameter)) +
  geom_histogram(bins=30, color="black", fill="grey", boundary = 0) + 
  labs(title = "Distribution of Pipe Diameter", 
       x = "Diameter(m)", 
       y = "Count")

# Box Plot for pipe diameter by conditions
b_pd <- ggplot(SW_data, aes(x=diameter, y = factor(structural_condition), fill = factor(structural_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Diameter by Structual Condition",
       fill = "Structural Condition",
          y = " ",
          x = "Diameter(m)") +
  scale_fill_discrete(labels = c("Good", "Poor"))

b_pd2 <- ggplot(SW_data, aes(x=diameter, y = factor(service_condition), fill = factor(service_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Diameter by Service Condition",
       fill = "Service Condition",
          y = " ",
          x = "Diameter(m)") +
  scale_fill_discrete(labels = c("Good", "Poor"))

ggarrange(h_pd, # first row with histogram plot
          ggarrange(b_pd, b_pd2, ncol=2), # second row with two box plots
          nrow = 2)
```

## 3.4 pipe length(m)
```{r}
# Histogram plot
h_pl <- ggplot(SW_data, aes(pipe_length)) +
  geom_histogram(bins=30, boundary = 0,color="black", fill="grey") +
  labs(title = "Distribution of Pipe Length",
       x = "Pipe Length(m)",
       y = "Count")
# Box Plot for pipe diameter by conditions
b_pl <- ggplot(SW_data, aes(x=pipe_length, y = factor(structural_condition), fill = factor(structural_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Pipe Length by Structual Condition",
       fill = "Structural Condition",
          y = " ",
          x = "Length(m)") +
  scale_fill_discrete(labels = c("Good", "Poor"))

b_pl2 <- ggplot(SW_data, aes(x=pipe_length, y = factor(service_condition), fill = factor(service_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Pipe Length by Service Condition",
       fill = "Service Condition",
          y = " ",
          x = "Length(m)") +
  scale_fill_discrete(labels = c("Good", "Poor"))

ggarrange(h_pl, # first row with histogram plot
          ggarrange(b_pl, b_pl2, ncol=2), # second row with two box plots
          nrow = 2)

```

## 3.5 Pipe Downstream Depth(m)
```{r}
# Histogram plot
h_d <- ggplot(SW_data, aes(downstream_depth)) +
  geom_histogram(bins=30, boundary = 0,color="black", fill="grey") +
  labs(title = "Distribution of Pipe Downstream Depth",
       x = "Pipe Depth(m)",
       y = "Count")
# Box Plot for pipe diameter by conditions
b_d <- ggplot(SW_data, aes(x=downstream_depth, y = factor(structural_condition), fill = factor(structural_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Downstream Depth by Structual Condition",
       fill = "Structural Condition",
          y = " ",
          x = "Depth(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 10))

b_d2 <- ggplot(SW_data, aes(x=downstream_depth, y = factor(service_condition), fill = factor(service_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Downstream Depth by Service Condition",
       fill = "Service Condition",
          y = " ",
          x = "Depth(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 10))

ggarrange(h_d, # first row with histogram plot
          ggarrange(b_d, b_d2, ncol=2), # second row with two box plots
          nrow = 2)
```

## 3.6 Pipe Upstream Depth(m)
```{r}
# Histogram plot
h_ud <- ggplot(SW_data, aes(upstream_depth)) +
  geom_histogram(bins=30, boundary = 0,color="black", fill="grey") +
  labs(title = "Distribution of Pipe Upstream Depth",
       x = "Pipe Depth(m)",
       y = "Count")
# Box Plot for pipe diameter by conditions
b_ud <- ggplot(SW_data, aes(x=upstream_depth, y = factor(structural_condition), fill = factor(structural_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Upstream Depth by Structual Condition",
       fill = "Structural Condition",
          y = " ",
          x = "Depth(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 10))

b_ud2 <- ggplot(SW_data, aes(x=upstream_depth, y = factor(service_condition), fill = factor(service_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Upstream Depth by Service Condition",
       fill = "Service Condition",
          y = " ",
          x = "Depth(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 10))

ggarrange(h_ud, # first row with histogram plot
          ggarrange(b_ud, b_ud2, ncol=2), # second row with two box plots
          nrow = 2)
```

## 3.7 Pipe Downstream Invert Level (m)
```{r}
SW_data %>%
  group_by(downstream_invert) %>%
  summarise(n = n())
# Histogram plot
h_dil <- ggplot(SW_data, aes(downstream_invert)) +
  geom_histogram(bins=30, boundary = 0,color="black", fill="grey") +
  labs(title = "Distribution of Pipe Downstream Invert Level",
       x = "Invert Level(m)",
       y = "Count")
# Box Plot for pipe diameter by conditions
b_dil <- ggplot(SW_data, aes(x=downstream_invert, y = factor(structural_condition), fill = factor(structural_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Downstream Invert Level by Structual Condition",
       fill = "Structural Condition",
          y = " ",
          x = "Invert Level(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 8))

b_dil2 <- ggplot(SW_data, aes(x=downstream_invert, y = factor(service_condition), fill = factor(service_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Downstream Invert Level by Service Condition",
       fill = "Service Condition",
          y = " ",
          x = "Invert Level(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 8))

ggarrange(h_dil, # first row with histogram plot
          ggarrange(b_dil, b_dil2, ncol=2), # second row with two box plots
          nrow = 2)
```

## 3.8 Pipe Upstream Invert Level (m)
```{r}
SW_data %>%
  group_by(upstream_invert) %>%
  summarise(n = n())
# Histogram plot
h_uil <- ggplot(SW_data, aes(upstream_invert)) +
  geom_histogram(bins=30, boundary = 0,color="black", fill="grey") +
  labs(title = "Distribution of Pipe Upstream Invert Level",
       x = "Invert Level(m)",
       y = "Count")
# Box Plot for pipe diameter by conditions
b_uil <- ggplot(SW_data, aes(x=upstream_invert, y = factor(structural_condition), fill = factor(structural_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Upstream Invert Level by Structual Condition",
       fill = "Structural Condition",
          y = " ",
          x = "Invert Level(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 8))

b_uil2 <- ggplot(SW_data, aes(x=upstream_invert, y = factor(service_condition), fill = factor(service_condition))) +
  geom_boxplot(alpha=0.4) +
  labs(title= "Upstream Invert Level by Service Condition",
       fill = "Service Condition",
          y = " ",
          x = "Invert Level(m)") +
  scale_fill_discrete(labels = c("Good", "Bad")) +
  theme(plot.title = element_text(size = 8))

ggarrange(h_uil, # first row with histogram plot
          ggarrange(b_uil, b_uil2, ncol=2), # second row with two box plots
          nrow = 2)
```


## 3.9 Other categorical variables
```{r}
# Pipe Asset Type
SW_data %>% 
    count(type = factor(asset_type)) %>% 
    mutate(pct = prop.table(n)) %>% 
    ggplot(aes(x = type, y = pct, label = scales::percent(pct))) + 
    geom_col(position = 'dodge', fill="grey", color="black") + 
    geom_text(position = position_dodge(width = .9),    # move to center of bars
              vjust = -0.5,    # nudge above top of bar
              size = 3) + 
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(guide = guide_axis(angle=45)) +
    labs(title = "Distribution of Asset Type",
         x = "Pipe Type",
         y = "Percentage")

# Pipe Shape
SW_data %>% 
    count(shape = factor(pipe_shape)) %>% 
    mutate(pct = prop.table(n)) %>% 
    ggplot(aes(x = shape, y = pct, label = scales::percent(pct))) + 
    geom_col(position = 'dodge', fill="grey", color="black") + 
    geom_text(position = position_dodge(width = .9),    # move to center of bars
              vjust = -0.5,    # nudge above top of bar
              size = 3) + 
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(guide = guide_axis(angle=45)) +
    labs(title = "Distribution of Pipe Shape",
         x = "Pipe Shape",
         y = "Percentage")

# Pipe catchment receiving environment
SW_data %>% 
    count(cre = factor(CRE)) %>% 
    mutate(pct = prop.table(n)) %>% 
    ggplot(aes(x = cre, y = pct, label = scales::percent(pct))) + 
    geom_col(position = 'dodge', fill="grey", color="black") + 
    geom_text(position = position_dodge(width = .9),    # move to center of bars
              vjust = 1,    # nudge inside the bar
              size = 3) + 
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(guide = guide_axis(angle=45)) +
    labs(title = "Distribution of Catchment Receiving Environment (CRE)",
         x = "Pipe CRE",
         y = "Percentage")

# Pipe Operational Ares
SW_data %>% 
    count(ope = factor(oper_area)) %>% 
    mutate(pct = prop.table(n)) %>% 
    ggplot(aes(x = ope, y = pct, label = scales::percent(pct))) + 
    geom_col(position = 'dodge', fill="grey", color="black") + 
    geom_text(position = position_dodge(width = .9),    # move to center of bars
              vjust = -0.5,    # nudge above top of bar
              size = 3) + 
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(guide = guide_axis(angle=45)) +
    labs(title = "Distribution of Pipe Operational Area",
         x = "Area",
         y = "Percentage")

# Pipe Local Boards
SW_data %>% 
    count(board = factor(local_board)) %>% 
    mutate(pct = prop.table(n)) %>% 
    ggplot(aes(x = board, y = pct, label = scales::percent(pct))) + 
    geom_col(position = 'dodge', fill="grey", color="black") + 
    geom_text(position = position_dodge(width = .9),    # move to center of bars
              vjust = -0.5,    # nudge above top of bar
              size = 3) + 
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(guide = guide_axis(angle=45)) +
    labs(title = "Distribution of Pipe Local Boards",
         x = "Surburbs",
         y = "Percentage")

```

# Correlation matrix 
```{r}
# Create a new dataframe only contains numeric variables
num_col <- SW_data[, c(1,3:6,8:9)]
# Convert two response variables from strings to numeric values
#num_col$structural_condition <- as.numeric(num_col$structural_condition)
#num_col$service_condition <- as.numeric(num_col$service_condition)
```
The chart.Correlation function is a shortcut to create a correlation plot in R with histograms, density functions, smoothed regression lines and correlation coefficients with the corresponding significance levels (if no stars, the variable is not statistically significant, while one, two and three stars mean that the corresponding variable is significant at 10%, 5% and 1% levels, respectively).
```{r}
chart.Correlation(num_col, histogram = TRUE, method = "pearson")
```
The corrplot.mixed function allows drawing correlation matrix with mixed visualization methods like circle, square, number, etc. The number shows coefficients with different color, and the area of circles indicate the absolute value of corresponding correlation coefficients.
```{r}
# correlation plot
corrplot.mixed(cor(num_col),
               lower = "number",
               upper = "circle",
               tl.col = "black",
               addgrid.col = "black",
               tl.cex = 0.6)
```

# 4. Variables Selection
```{r}
# response variable: structural condition
# Data Preparation
SW <- subset(SW_data,
                select = c(years, material, diameter,
                           pipe_length, downstream_depth,
                           upstream_depth, pipe_shape,
                           downstream_invert,
                           upstream_invert, 
                           oper_area,
                           local_board,
                           structural_condition,
                           service_condition))
# Export the clean data frame to .csv
write.csv(SW, 
          "C:/Users/fan100199/OneDrive - Mott MacDonald/Desktop/MottMac/Data_sets\\readytogo.csv",
          row.names = FALSE)




```


